Isolator: accurate and consistent analysis of isoform-level expression in RNA-Seq experiments

### 2016-09-23 paper.tex
Try to solidify discussions of consistency/stability.

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex or similar sets of isoforms. As a result, estimates of
    isoform abundance carry far more uncertainty than those made at the gene
    level. When confronted with this uncertainty, commonly used methods produce
    estimates that are unstable---small perturbations in the data often produce
    dramatically different results, confounding downstream analysis. We
    introduce a new method, Isolator, which analyzes all samples in an
    experiment in unison using a simple Bayesian hierarchical model. Combined
    with aggressive bias correction, it produces estimates that are
    simultaneously accurate and stable. In a comprehensive comparison of
    accuracy and stability, we show that this property is unique to Isolator. We
    further demonstrate that the approach of modeling an entire experiment
    enables new analyses, which we demonstrate by examining splicing
    monotonicity across several time points in the development of human
    cardiomyocyte cells.
\end{abstract}

@@

We built on this work, implementing a new method called Isolator. Though
it retains the same underlying model of RNA-Seq, we approach the problem
from the perspective that isoform-level estimates possess a degree of
uncertainty that renders maximum likelihood point-estimates alone a
fundamentally inadequate solution to the problem. As the purpose of gene
expression studies is to compare expression between multiple samples,
our goal, beyond maximizing a narrow definition of accuracy, is to
provide a tool that effectively accounts for this uncertainty in a
coherent and reliable manner.

@@

The convention for evaluating the accuracy of RNA-Seq analysis methods has
consisted chiefly of comparing point estimate accuracy in simulations,
supplemented with comparisons to qPCR or other technologies. Tuning methods
exclusively to maximize accuracy on simulated data risks solving only an
idealized mathematical model of RNA-Seq while disregarding noise, bias, and
nonuniformity. As no gold standard for whole-transcriptome isoform-level
estimation exists, we adopt a multifaceted approach, focusing on measuring
the stability or consistency of estimates, along with accuracy.

Comparisons of gene or transcript expression are sensitive to how one chooses to
measure similarity. Because gene expression typically varies across many orders
of magnitude, measurements like the Pearson correlation, $L_1$ or $L_2$
distance, or root-mean-square error are often dominated by the most highly
expressed genes. Recognizing this, a popular alternative is Spearman's rank
correlation, which conversely can overemphasize the majority of transcripts,
which have no, or very low expression. These issues are discussed in depth by
Lovell et al \cite{Lovell:2015il}, who propose measurements of proportionality
as a more meaningful metric. We adopt their ``proportionality correlation'',
defined as $\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) /
(\text{var}(\log(x)) + \text{var}(\log(y)))$, where is $\text{cov}(x)$ and
$\text{var}(x)$ give the covariance and variance of $x$, respectively. Similar
to other measures of correlation, proportionality correlation varies from -1 to
1, with 1 indicating perfectly proportionality, and -1 perfect reciprocality.
Zeros are accounted for by converting all measurements to transcripts per
million (TPM) and adding 0.1. Results were found to be fairly insensitive to the
precise additive constant used here (Supplementary Section
\ref{additive-constant}). Though we believe proportionality correlation is a
superior approach, much of this analysis was also conducted using Spearman's
rank correlation in Supplementary Section \ref{spearman-reanalysis}, and largely
agrees with the results shown here.

@@

RSEM was used to produce both maximum likelihood and posterior mean estimates,
which we label ``RSEM/ML'' and ``RSEM/PM'', respectively.  This is not an
exhaustive account of such methods, but does represent a wide variety of popular
approaches. When possible, the same read alignments were used for each method,
which notably excludes the alignment-free methods Sailfish, Salmon, and
Kallisto (Supplementary Section \ref{analysis}).

@@

\caption{
Benchmarks measuring the accuracy and consistency of estimates made from
Sequencing Quality Control (SEQC) dataset using a variety of methods.
All numbers are reported in proportionality correlation.
(a) correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.
(b) correlation between gene level quantification
of 18,353 genes using PrimePCR qPCR and RNA-Seq quantification.
(c) correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.
(d) correlation between transcript-level estimates
for the mixed SEQC samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D. Sample C consists of
$3/4$ A and $1/4$ B and sample D is $1/4$ A and $3/4$ B. Consistent estimates of
transcript expression in A, B, C, and D should agree with these mixture
proportions.}
\end{figure*}

\subsection{Estimation consistency}\label{estimation-consistency}

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. This design allows us to directly
measure expression consistency. If we estimate transcript abundance for each
sample, producing estimates $a, b, c, d$, then for these estimates to be in
agreement we should expect $c$ to be approximated by $0.25a + 0.75b$, and
similarly $d$ by $0.75a + 0.25b$. To capture the notion of consistency, we
measured the closeness of this approximation using proportionality correlation
(Figure \ref{table:seqctranscripts}).

@@

We further examined the question of consistency by comparing the same
samples sequenced on different flow cells and at different sites, again
using data from the SEQC. We used 10 samples from the larger SEQC
dataset, each consisting of a single lane from two separate
flowcells sequenced at five different sites, all on Illumina HiSeq 2000
instruments, and compared pairwise agreement between estimates from these
samples (Figure \ref{fig:batcheffects}a). In the interest of a more fair
comparison, Isolator was run on each sample in isolation without sharing
any information.

@@

To assess the suitability of Isolator we have compared it to large number of
alternative methods using a wide variety of benchmarks, with a particular focus
on estimation consistency or stability. Data generated by the SEQC is ideally
suited to examine the question consistency, and indeed our data agrees with
their results. The SEQC compared Cufflinks and BitSeq and found BitSeq to
produce more consistent estimates. Here we control for alignment method by using
the same alignments for every method (excluding the alignment-free methods
Salmon, Sailfish, and Kallisto), and compare a wider variety methods. In this
broader context it becomes clear that this separation is a consequence of the
choice in estimators. Unmoderated maximum likelihood methods, though often
accurate, frequently demonstrate lower consistency than methods that use MCMC
sampling and report posterior means or medians. This is underappreciated but
should be unsurprising.  Because of the ambiguity of sequencing data, the
likelihood functions for relative expression are often near plateaus. Maximum
likelihood estimates are thus precarious compared to posterior mean
estimates. Neither estimate is inherently more correct, and robust analysis
should make use of the full posterior distribution, but we argue that posterior
mean or median estimates are more useful in practice because of their greater
stability.

Excluding Isolator, methods that report posterior mean estimates show seemingly
poor accuracy. This is largely driven by low-expression transcripts. A
transcript with no reads has a maximum likelihood expression of zero, but its
likelihood is positive over a range of non-zero values, so the posterior mean
must necessarily also be positive. With a flat, or ``non-informative,'' prior,
expression values for these low-expression transcripts are often quite high and
highly determined by transcript length.  Isolator avoids this problem by using a
model with informative priors: namely encoding assumptions that most transcripts
have low expression and that expression and splicing tend to be similar between
samples. In this way, it is able to avoid what seems like a tradeoff and
produces estimates that are both consistent and accurate.

@@

In this paper we have described a method that effectively shares information
between samples in RNA-Seq experiments while rigorously quantifying uncertainty,
leading to more accurate results across a wide range of benchmarks. We believe
this work indicates a natural direction towards more reliable and powerful
studies of transcriptomics, which future work can follow by scaling this
methodology to larger numbers of samples, and accounting for more forms of
uncertainty and technical effects.

>>>>

\begin{abstract}
    While RNA-Seq has enabled great progress towards the goal of wide-scale
    isoform-level mRNA quantification, short reads have limitations when
    resolving complex or similar sets of isoforms. As a result, estimates of
    isoform abundance carry far more uncertainty than those made at the gene
    level. When confronted with this uncertainty, commonly used methods produce
    estimates that are often high-variance---small perturbations in the data
    often produce dramatically different results, confounding downstream
    analysis. We introduce a new method, Isolator, which analyzes all samples in
    an experiment in unison using a simple Bayesian hierarchical model. Combined
    with aggressive bias correction, it produces estimates that are
    simultaneously accurate and stable. In a comprehensive comparison of
    accuracy and variance, we show that this property is unique to Isolator. We
    further demonstrate that the approach of modeling an entire experiment
    enables new analyses, which we demonstrate by examining splicing
    monotonicity across several time points in the development of human
    cardiomyocyte cells.
\end{abstract}

@@

We built on this work, implementing a new method called Isolator. Though it
retains the same underlying model of RNA-Seq, we approach the problem from the
perspective that isoform-level estimates possess a degree of uncertainty that
renders unregularized maximum likelihood point-estimates alone a fundamentally
inadequate solution to the problem. As the purpose of gene expression studies is
to compare expression between multiple samples, our goal, beyond maximizing a
narrow definition of accuracy, is to provide a tool that effectively accounts
for this uncertainty in a coherent and reliable manner.

@@

The convention for evaluating the accuracy of RNA-Seq analysis methods has
consisted chiefly of comparing point estimate accuracy in simulations,
supplemented with comparisons to qPCR or other technologies. Tuning methods
exclusively to maximize accuracy on simulated data risks solving only an
idealized mathematical model of RNA-Seq while disregarding noise, bias, and
nonuniformity. As no gold standard for whole-transcriptome isoform-level
estimation exists, we adopt a multifaceted approach, focusing on measuring
the variance of estimates, along with overall accuracy.

Comparisons of gene or transcript expression are sensitive to how one chooses to
measure similarity. Because gene expression typically varies across many orders
of magnitude, measurements like the Pearson correlation, $L_1$ or $L_2$
distance, or root-mean-square error are often dominated by the most highly
expressed genes. Recognizing this, a popular alternative is Spearman's rank
correlation, which conversely can overemphasize the majority of transcripts,
which have no, or very low expression. These issues are discussed in depth by
Lovell et al \cite{Lovell:2015il}, who propose measurements of proportionality
as a more meaningful metric. We adopt their ``proportionality correlation'',
defined as $\text{pcor}(x, y) = 2\text{cov}(\log(x), \log(y)) /
(\text{var}(\log(x)) + \text{var}(\log(y)))$, where is $\text{cov}(x)$ and
$\text{var}(x)$ give the covariance and variance of $x$, respectively. Similar
to other measures of correlation, proportionality correlation varies from -1 to
1, with 1 indicating perfect proportionality, and -1 perfect reciprocality.
Zeros are accounted for by converting all measurements to transcripts per
million (TPM) and adding 0.1. (Results were found to be fairly insensitive to the
precise additive constant used here (Supplementary Section
\ref{additive-constant})). Though we believe proportionality correlation is a
superior approach, much of this analysis was also conducted using Spearman's
rank correlation in Supplementary Section \ref{spearman-reanalysis}, and largely
agrees with the results shown here.

@@

RSEM was used to produce both maximum likelihood and posterior mean estimates,
which we label ``RSEM/ML'' and ``RSEM/PM'', respectively. This is not an
exhaustive account of such methods, but does represent a wide variety of popular
approaches. The same read alignments were used for all methods other Sailfish,
Salmon, and Kallisto, which are ``alignment-free'', using built-in
approximations to full alignment (Supplementary Section \ref{analysis}).

@@

\caption{
Benchmarks measuring the accuracy and coherence of estimates made from
Sequencing Quality Control (SEQC) dataset using a variety of methods.
All numbers are reported in proportionality correlation.
(a) correlation between gene level quantification
of 806 genes using TaqMan qPCR and RNA-Seq quantification.
(b) correlation between gene level quantification
of 18,353 genes using PrimePCR qPCR and RNA-Seq quantification.
(c) correlation between known proportions of 92
ERCC spike-in controls and RNA-Seq quantification.
(d) correlation between transcript-level estimates
for the mixed SEQC samples C and D and weighted averages of estimates for A
and B, corresponding to the mixture proportions for C and D. Sample C consists of
$3/4$ A and $1/4$ B and sample D is $1/4$ A and $3/4$ B. Consistent estimates of
transcript expression in A, B, C, and D should agree with these mixture
proportions.}
\end{figure*}

\subsection{Estimate coherence}\label{estimation-coherence}

The four SEQC samples consist of two commercially available reference RNA
samples, labeled $A$ and $B$, and two synthetic samples formed by mixing $A$ and
$B$ in specific proportions. Sample $C$ is composed of 75\% $A$ and 25\% $B$,
while sample $D$ is 25\% $A$ and 75\% $B$. If we estimate transcript abundance
for each sample, producing estimates $a, b, c, d$, then for these estimates to
be in agreement (``coherent'', in our terminology) we should expect $c \appprox
0.25a + 0.75b$, and similarly $d \approx 0.75a + 0.25b$. In other words, mixing
RNA then estimating transcript abundance should be approximated by estimating
transcript abundance then mixing estimates.

To capture the notion of coherence, we measured the closeness of this
approximation using proportionality correlation (Figure
\ref{table:seqctranscripts}).

@@

We further examined the question of estimate variance by comparing the same
samples sequenced on different flow cells and at different sites, again using
data from the SEQC. We used 10 samples from the larger SEQC dataset, each
consisting of a single lane from two separate flowcells sequenced at five
different sites, all on Illumina HiSeq 2000 instruments, and compared pairwise
agreement between estimates from these samples (Figure \ref{fig:batcheffects}a).
In the interest of a more fair comparison, Isolator was run on each sample in
isolation without sharing any information.

@@

To assess the suitability of Isolator we have compared it to large number of
alternative methods using a wide variety of benchmarks, with a particular focus
on estimation variance and coherence. Data generated by the SEQC is ideally
suited to examine this question, and indeed our analysis agrees with their
results. The SEQC compared Cufflinks and BitSeq and found BitSeq to produce more
consistent, less variable estimates. We took a closer look at this phenomenon,
controlling for alignment method by using the same set of alignments for every
method (excluding the alignment-free methods Salmon, Sailfish, and Kallisto),
and comparing a wider variety of methods. In this broader context it becomes
clear that this separation is a consequence of the choice in estimators.
Unmoderated maximum likelihood methods, though often accurate (i.e. low bias),
frequently demonstrate less consistency (i.e. high variance) than methods that
use MCMC sampling and report posterior means or medians. This is
underappreciated but should be unsurprising.  Because of the ambiguity of
sequencing data, the likelihood functions for relative expression are often
nearly plateaus. Maximum likelihood estimates are thus precarious compared to
posterior mean estimates. Neither estimate is inherently more correct, and
robust analysis should make use of the full posterior distribution, but we argue
that posterior mean or median estimates are more useful in practice because they
tend to have lower variance.

Excluding Isolator, methods that report posterior mean estimates show seemingly
poor accuracy. This is largely driven by low-expression transcripts. A
transcript with no reads has a maximum likelihood expression of zero, but its
likelihood is positive over a range of non-zero values, so the posterior mean
must necessarily also be positive. With a flat, or ``non-informative,'' prior,
expression values for these low-expression transcripts are often quite high and
highly determined by transcript length. Isolator avoids this problem by using a
model with informative priors: namely encoding assumptions that most transcripts
have low expression and that expression and splicing tend to be similar between
samples. In this way, it is able to avoid what seems like a tradeoff and
produces estimates that are both relatively low bias and low variance.

@@

In this paper we have described a method that effectively shares information
between samples in RNA-Seq experiments while rigorously quantifying uncertainty,
leading to more accurate results across a wide range of benchmarks. We believe
this work indicates a natural direction towards more reliable and powerful
studies of transcriptomics, which future work can follow by scaling this
methodology to larger numbers of samples, allowing for more flexible models, and
accounting for more forms of uncertainty and technical effects.
